COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-03 paper/bare_conf.tex
Minor rewording of abstract

\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
Florian Rathgeber\IEEEauthorrefmark{1},
Ana Lucia Varbanescu\IEEEauthorrefmark{2}, 
Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
David A. Ham\IEEEauthorrefmark{4},
J. Ramanujam\IEEEauthorrefmark{3},
and
Paul H. J. Kelly\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
\IEEEauthorrefmark{4}Imperial College London, Department of Computing and Department of Mathematics,\\
\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$d.ham\}@imperial.ac.uk,\\
\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
\IEEEauthorrefmark{3}jxr@ece.lsu.edu
}
}

@@

\begin{abstract}
%\boldmath
Numerical solution of partial differential equations using the finite element method is one of the key applications of high performance computing. The local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain. Since the domain size can be huge, executing efficient kernels is fundamental. Their optimization is, however, a challenging issue. Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. We present, therefore, the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Experiments using a suite of real-world examples show that significant performance improvement is achieved.
\end{abstract}

@@

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to reach close-to-peak performance. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-results}, in which manual BLAS-based implementations are evaluated, the small dimension of the involved matrices and loss in data locality can limit the performance gain, if any.

@@

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contribution of this paper is threefold
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. polyhedral) compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel machines, a Sandy Bridge and the Phi.
\end{itemize}

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time become the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have. First, they can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the default loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion) and we ignore the presence of \texttt{det*W3[i]}, we can notice that the computation of the stiffness matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the stiffness matrix computation is always reducible to this kind of form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when using unroll-and-jam, with $f$ being the unroll-and-jam factor.

@@

The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$, in the following). We efficiently restore it with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. This high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead. Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several Finite-Volume-based applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced. Their implementation is based on analysis and transformation of the kernel's AST. When $op$-$vect$ is selected, the compiler outputs proper AVX (or LRBNi) intrinsics code. Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers are available to carry out the computation of the stiffness matrix.

@@

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order. These are used to estimate unroll-and-jam and split factors when, respectively, $op$-$vect$ and $split$ are used. The $n\_consts$ parameter includes temporary registers to carry out computations. If a factor is 0, then the corresponding transformation is not applied. The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16).

@@

\subsection{Comparison with Fenics Form Compiler's built-in Optimizations}
We have modified the Fenics Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE. Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued columns in basis function arrays. The FEniCS' loop-invariant code motion is, however, definitely different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest. Thus, depending on the way expansion is performed, operation count does not necessarily decrease. On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code. This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which affect vectorization because of the need for gather operations. The price to pay is that only some of the zero-columns will be removed.

@@

In this paper we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly. In this context, to the best of our knowledge, this is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization. Assembly kernels have peculiar characteristics. Their iteration space is usually very small, with the size depending on aspects like the degree of accuracy one wants to reach (polynomial order of the method) and the mesh discretization employed. The data space, in terms of number of matrices and scalar, grows proportionally with the complexity of the finite element problem. COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is simply too superficial if close-to-peak performance needs to be reached. The various optimizations overcome limitations of current vendor and research compilers. The exploitation of domain-knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms. Further work include a comprehensive study about feasibility and constraints on transforming the kernel into a sequence of BLAS calls. COFFEE supports all of the problems expressible in Firedrake, and it is already integrated with this framework.

>>>>

\author{\IEEEauthorblockN{Fabio Luporini\IEEEauthorrefmark{1},
Ana Lucia Varbanescu\IEEEauthorrefmark{2},
Florian Rathgeber\IEEEauthorrefmark{1},
Gheorghe-Teodor Bercea\IEEEauthorrefmark{1}, \\
J. Ramanujam\IEEEauthorrefmark{3},
David A. Ham\IEEEauthorrefmark{4},
and
Paul H. J. Kelly\IEEEauthorrefmark{1}}
\IEEEauthorblockA{\IEEEauthorrefmark{1}Imperial College London, Department of Computing,\\
\IEEEauthorrefmark{4}Imperial College London, Department of Computing and Department of Mathematics,\\
\IEEEauthorrefmark{2}TU Delft, Faculty of Engineering, Mathematics and Computer Science (EWI)\\
\IEEEauthorrefmark{1}\{f.luporini12$|$f.rathgeber10$|$gheorghe-teodor.bercea08$|$p.kelly$|$david.ham\}@imperial.ac.uk,\\
\IEEEauthorrefmark{2}a.l.varbanescu@uva.nl,
\IEEEauthorrefmark{3}jxr@ece.lsu.edu
}
}

@@

\begin{abstract}
%\boldmath
The numerical solution of partial differential equations using the finite element method is a key application in scientifc computing. Local assembly is its characteristic operation, which executes a problem-specific kernel for each element in the discretized problem domain. Since the domain size can be huge, executing efficient kernels is fundamental. Their optimization is, however, a challenging issue. Even though affine loop nests are generally present, the short trip counts and the complexity of mathematical expressions make it hard to determine a single or unique sequence of successful transformations. We present the design and systematic evaluation of COFFEE, a domain-specific compiler for local assembly kernels. COFFEE manipulates abstract syntax trees by introducing composable optimizations aimed at improving instruction-level parallelism, especially SIMD vectorization, and register locality. It then generates C code including vector intrinsics. Experiments using a suite of real-world examples show that significant performance improvement is achieved.
\end{abstract}

@@

Achieving high-performance is non-trivial. The complexity of the mathematical expressions, often characterized by a large number of operations on constants and small matrices, makes it hard to determine a single or specific sequence of transformations that is successfully applicable to all problems. The variation in loop trip counts and their typically small value, which prevents the adoption of standard techniques for cache locality, further exacerbates the issue. A compiler-based approach is, therefore, the only reasonable option to obtain close-to-peak performance in a wide range of different local assembly kernels. Optimizations like generalized loop-invariant code motion, vector-register tiling, and expression splitting, as well as their composition, are essential, but they are not supported by state-of-the-art polyhedral and vendor compilers. BLAS routines could be theoretically employed, although a fairly complicated control- and data-flow analysis would be required to automate identification and extraction of matrix-matrix multiplies. In addition, as detailed in Section~\ref{sec:perf-results}, in which manual BLAS-based implementations are evaluated, the small dimension of the involved matrices and loss in data locality can limit the performance gain, if any.

@@

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 1.5$\times$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as generalized loop-invariant code motion, padding, and data alignment), and up to 4.44$\times$ over original kernels. The contributions of this paper are threefold
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely local assembly in the context of the finite element method. Our approach, which also exploits domain knowledge, goes beyond the limits of both vendor and research compilers.
\item Design and implementation of a compiler that automates the proposed code transformations for any problems expressible in Firedrake.
\item Systematic evaluation using a suite of examples of real-world importance, and evidence of significant performance improvements on two Intel machines, a Sandy Bridge and the Phi.
\end{itemize}

@@

\section{Code Transformations}
\label{sec:code-transf}
The code transformations presented in this section are applicable to all finite element problems that can be formulated in Firedrake. One peculiar characteristic is that they all aim at improving the run-time of the \texttt{ijk} loop nest, where the numerical integration takes place. In rare, yet important cases, however, this might not be sufficient to achieve notable speed-ups. Generalized loop-invariant code motion, which is described in Section~\ref{sec:code-transf-licm}, is a fundamental optimization that allows pre-computation of invariant sub-expressions. It is worth noting there are circumstances in which the amount of lifted terms, either at the level of an outer loop or even completely outside of the loop nest, is so big (thousands of operations, hundreds of temporaries) that their execution time becomes the dominant factor of the overall local assembly run-time. In these cases, two challenging optimizations are common sub-expression elimination and effective vectorization (superword level parallelism~\cite{SLP} would not be of help, because invariant sub-expressions lack, in general, structure). Given its inherent complexity, a comprehensive study of this problem is left as further work.

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Generalized Loop-invariant Code Motion}
\label{sec:code-transf-licm}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $intel$ and $gnu$ compilers have. First, they can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted in the loop order \texttt{ijk}. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along its fastest varying dimension. This is implemented by introducing a temporary array per invariant sub-expression and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Model-driven Vector-register Tiling}
One notable problem of assembly kernels concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Since the kernel's working set fits the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, and unroll-and-jam, can be employed to deal with this problem. In COFFEE, these optimizations are supported either by means of explicit code transformations (interchange, unroll-and-jam) or indirectly by delegation to the compiler through standard pragmas (unroll). Tiling at the level of vector registers is an additional feature of COFFEE. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, a model-driven vector-register tiling strategy can be implemented. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} and we ignore the presence of the operation \texttt{det*W3[i]}, we can notice that the computation of the stiffness matrix is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\
  y \in B'' \subseteq B}}
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all basis functions (or temporary variables, e.g. \texttt{LI$\_$0}) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Regardless of the specific input problem, and by abstracting from the presence of all variables independent of both $j$ and $k$, the stiffness matrix computation is always reducible to this form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers, which represent a 4$\times$4 tile, assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gnu$ and $intel$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation. Besides a larger number of cache accesses, it needs to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

The storage layout of $A$, however, is incorrect after the application of this outer-product-based vectorization ($op$-$vect$, in the following). It can be efficiently restored with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the \texttt{ijk} loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

@@

Firedrake provides users with the Unified Form Language to write problems in a notation resembling mathematical equations. At run-time, this high-level specification is translated by the Fenics Form Compiler~\cite{FFC-Compiler} into an abstract syntax tree (AST) representation of one or more finite element assembly kernels. ASTs are then passed to COFFEE, which is capable of applying the transformations described in Section~\ref{sec:code-transf}. The output of COFFEE is C code, which is eventually provided to PyOP2~\cite{pyop2isc}, where just-in-time compilation and parallel execution over the unstructured mesh take place. Because of the large number of (possibly parametric) transformations, COFFEE needs a mechanism to select the most suitable optimization strategy for a given problem. Auto-tuning might be used, although at the moment we avoid it to minimize the run-time overhead (the generation of the AST occurs at run-time as soon as problem-specific data are available). Our optimization strategy, based on heuristics and a simple cost model, is described in the following, along with an overview of the compiler.

The compiler structure is outlined in Figure~\ref{algo:PyOP2Compiler}. Initially, an AST is inspected, looking for the presence of iteration spaces and domain-specific information provided by the higher layer. If the kernel lacks an iteration space, then so-called inter-kernel vectorization, in which the non-affine loop over mesh elements is vectorized, can be applied. This feature, currently under development, has been proved to be useful in several finite volume applications~\cite{inter-kernel-vect}. The second transformation step is applied if the backend is a manycore machine, like a GPU: the compiler tries to extract parallelism from inside the kernel, by partitioning loop iterations among different threads, if these are found to be independent~\cite{pyop2isc}. Then, an ordered sequence of optimization steps are executed. Application of $licm$ must precede padding and data alignment, due to the introduction of temporary arrays. Based on a cost model, loop interchange, $split$, and $op$-$vect$ may be introduced. Their implementation is based on analysis and transformation of the kernel's AST. When $op$-$vect$ is selected, the compiler outputs proper AVX or LRBNi intrinsics code. Any possible corner cases are handled: for example, if $op$-$vect$ is to be applied, but the size of the iteration space is not a multiple of the vector length, then a reminder loop, amenable to auto-vectorization, is inserted.

All loops are interchangeable provided that temporaries are introduced if the nest is not perfect. For the employed storage layout, the loop permutations \texttt{ijk} and \texttt{ikj} are likely to maximize performance. Conceptually, this is motivated by the fact that if the $i$ loop were in an inner position, then a significantly higher number of load instructions would be required every iteration; experiments showed that the performance loss is greater than the gain due to the possibility of accumulating increments in a register, rather than in memory, along the $i$ loop. The choice between \texttt{ijk} and \texttt{ikj} depends on the number of load instructions that can be hoisted out of the innermost dimension. Our compiler chooses, as outer, the loop along which the number of invariant loads is smaller so that more registers remain available to carry out the computation of the stiffness matrix.

@@

The cost model is shown in Figure~\ref{algo:applyCostModel}. It takes into account the number of available logical vector registers, $n\_regs$, and the number of unique variables accessed: $n\_consts$ counts variables independent of both $j$ and $k$ loops and temporary registers, $n\_outer\_arrays$ counts $j$-dependent variables, and $n\_inner\_arrays$ counts $k$-dependent variables, assuming the \texttt{ijk} loop order. These values are used to estimate unroll-and-jam and split factors for $op$-$vect$ and $split$. If a factor is 0, then the corresponding transformation is not applied. The $split$ transformation is triggered whenever the number of hoistable terms is larger than the available registers along the outer dimension (lines 3-8), which is approximated as half of the total (line 2). A split factor of $n$ means that the assembly expression should be ``cut'' into $n$ sub-expressions. Depending on the structure of the assembly expression, each sub-expression might end up accessing a different number of arrays; the cost model is simplified by assuming that all sub-expressions are of the same size. The unroll-and-jam factor for the $op$-$vect$ transformation is determined as a function of the available logical registers, i.e. those not used for storing hoisted terms (line 9-11). Finally, the profitability of loop interchange is evaluated (line 12-16).

@@

\subsection{Comparison with Fenics Form Compiler's built-in Optimizations}
We have modified the Fenics Form Compiler (FFC) to make it return an abstract syntax tree representation of a local assembly kernel, rather than plain C code, so as to enable code transformations in COFFEE. Besides Firedrake, FFC is used in the FEniCS project~\cite{Fenics}. In FEniCS, FFC can apply its own model-driven optimizations to local assembly kernels~\cite{quadrature1}, which mainly consist of loop-invariant code motion and elimination, at code generation time, of floating point operations involving zero-valued columns in basis function arrays. The FEniCS' loop-invariant code motion is, however, definitely different from COFFEE's, since it is based on expansion of arithmetic operations to identify terms that are invariant of the whole loop nest. Thus, depending on the way expansion is performed, operation count does not necessarily decrease. On the other hand, elimination of zero-valued columns, which are due to certain properties of the input finite element problem, has the effect of introducing indirection arrays in the generated code. This kind of optimization is currently under development in COFFEE, although it will differ from FEniCS' by avoiding non-contiguous memory accesses, which would affect vectorization because of the need for gather operations. The price to pay is that only some of the zero-columns will be removed.

@@

In this paper we have presented design, optimizations and systematic performance evaluation of COFFEE, a compiler for finite element local assembly. In this context, to the best of our knowledge, this is the first compiler oriented towards the introduction of low-level optimizations to maximize instruction-level parallelism, register locality and SIMD vectorization. Assembly kernels have peculiar characteristics. Their iteration space is usually very small, with the size depending on aspects like the degree of accuracy one wants to reach (polynomial order of the method) and the mesh discretization employed. The data space, in terms of number of matrices and scalar, grows proportionally with the complexity of the finite element problem. COFFEE has been developed taking into account all of these degrees of freedom, based on the idea that reducing the problem of local assembly optimization to a fixed sequence of transformations is way too superficial if close-to-peak performance needs to be reached. The various optimizations overcome limitations of current vendor and research compilers. The exploitation of domain-knowledge allows some of them to be particularly effective, as demonstrated by our experiments on two state-of-the-art Intel platforms. Further work include a comprehensive study about feasibility and constraints on transforming the kernel into a sequence of BLAS calls. COFFEE supports all of the problems expressible in Firedrake, and it is already integrated with this framework.
